\chapter{Introduction}
% TODO missing finance concepts

% Investment banks
% - Equity division
% - Delta-one trading

% TODO move finance notes from evernote to here

% Numerical methods
% - option pricing https://en.wikipedia.org/wiki/Binomial_options_pricing_model

% create-redeem process

% Financial producta
% - Options contracts
% - Cash equities
% - Convertible bonds
% total return swaps,
% ETFs, the exchange-traded fund (ETF), which combined the structure of a passive index fund with the trading characteristics of futures that were most attractive to active traders: intraday liquidity, the ability to go short and, with the help of a prime broker or margin account, the ability to obtain leverage A particular advantage of ETFs, as compared to index futures, was that it was considerably quicker and easier to create new products tracking different indices. As a result, the breadth of coverage of ETFs expanded rapidly to include not just country indices but indices based on sectors, regions, market capitalisation ranges, emerging markets and global benchmarks
% futures,
% forwards,
% certificates,
%  warrants
% fixed income

% risk analysis, where "delta" is the measurement of the sensitivity of a derivative to movements in the price of the underlying asset class or security

\section{Delta-one trading}
% TODO extract index fund from delta trading
Wwhy would an investor purchase an asset that provided economic exposure to an underlier as opposed to purchasing the underlier directly. The reason for this is quite simply that indices are themselves not tradeable products. An equity index is ultimately a number, whose value represents the weighted average price of a portfolio of securities. When an investor wishes to gain exposure to an index he/she must either:

- Purchase all of the underlying securities of the index in their respective weights and keep the composition up to date as weights change due to corporate actions, rebalances, etc.
- Purchase a product that replicates the index performance.

Designed as tools for long-term buy-and-hold strategies, index funds are not intended as trading tools and cannot be executed intraday. to fill this gap equity index futures was invented. Here investors could ...
% TODO

With commodity futures, the terms of the contract specify physical delivery of the underlying asset at maturity. For equity index futures, the delivery of the underlier is impossible and the settlement of a replicating stock portfolio is logistically impractical. As a result, index futures were settled via a cash payment between buyers and sellers, based on an objective reference price.

Passive index funds can be considered the first delta-one products
