\chapter{Preface} 
It has been said that the typical theorem in mathematics states that something you do not understand is equal to something else
you cannot compute. There is a kernel of truth in this joke, since the rigor required when doing mathematical research has found
is way into every nook and corner of every textbook and thereby to a large extend rendered them unreadable to all, except fairly
expert mathematicians; and even these usually disdain from reading such texts cover to cover. The value of mathematical rigor is
not in the question; but such works are condemned to remain works of reference, to be consulted, not read.

This is unfortunate since many people derive great happiness from fiddling with a little mathematics; just consider the many
people who vigorously engages the daily news paper Sudoku puzzel on their way to work. Such spending of time implies that many
people indeed still enjoys the pleasures of the mind. Thus the current crisis in science is to a large extend self inflicted,
modern scientist have done to little to convey what they are doing by hiding behind the often tutted vail of complexity.

However it would be foolish not to acknowledge the complexity of modern mathematical developments, such as Andrew Wiles incredible
proof  of Fermat's Last Theorem or Abel's remarkable proof of the in-solvability of general higher degree equations. Such gems are
usually only available to the person equipped with a sharp mind, a fairly developed bag of mathematical tricks and a certain
intellectual maturity. What I propose here is a book that will build  up enough knowledge as to show the proofs of the greatest
theorems in the history of mathematics;  from Hippocrates Quadrature of the Lune to Lindeman's proof that Pi is Trancendental. I
have sought to make the text completely self contained, so the reader will be taken from the proof that $a^2 + b^2 = c^2$ holds
true for the length of the sides in a right triangle, to the complex proof that if an integer $n$ is greater than 2, then the
equation $x^n + y^n = z^n$ has no solutions for non-zero integers $x$, $y$, and $z$. 

The mathematical mature person is right to ask if such an bold attempt is not futile and that perhaps this author is a bit naive.
Such a question is indeed prudent to ask, but for its answer I need only refer to the massive success of Richard P. Feynman'a
lectures of physics; in which the author successfully conveys such complex matters as Quantum mechanics and General relativity to
people with little other background than a high school education and a lust to learn a little physics. There goes a story that
Feynman was once asked by a Caltech faculty member to explain why spin 1/2 particles obey the Fermi-Dirac statistics. He nodded
and said , "I'll prepare a freshman lecture on it". But a few days later he returned and said. "You know, I couldn't do it, I
couldn't reduce it to freshman level. That means we really don't understand it." 

The text includes a number of exercises along with answers to every one (at least briefly). I am a firm believer in
learning-by-looking, also known as cheating, and thus I encourage the reader to give each problem his best shot, then check the
answer. If correct then move on to a more advanced problem, if not then study the provided solution and try to solve a similar
problem. If you are stuck then don't panic I have included an entire chapter devoted to the fine art of problem solving and
mathematical proof. The exercise are grouped as follows
\begin{itemize}
\item \textbf{Warmups} Problems that everyone should try to solve, even if they appear trivial. As the late great mathematician
George Polya (1887 - 1985) said  "The advanced reader who skips parts that appear too elementary may miss more than the less
advanced reader who skips parts that appear too complex". 
\item \textbf{Study Exercise} Exercises that will develop and widen the ideas from the text, many problems in this section
resembles university exam problems.
\item \textbf{Research Problems} Advanced and interesting problems that may or may not be solvable. These problems provides your
window to mathematical fame, but do not discourage if you fail as at least one person (the author) have failed here before you. 
\end{itemize}
The author will be interested to learn of any solution or headway made into the research problems as well as any simplification or clarification to the non-research ones. The main text will never refer to results developed in the exercises (a very common sin in mathematics), thus every vital result in this book is presented in entirety. This does however not mean that the exercises should be skipped, because in order to understand mathematics you have to do mathematics.

The subject presented here is not laid out in the usual chronological manner, instead I have sought to build up mathematics as a
logical progression of ever more complex ideas. Therefor instead of focusing on historical order I have strived instead to provide
the clearest proofs; so for example, instead of Oresme's original verbal proof of the divergence of the Harmonic Series I have
instead included a never and simpler algebraic proof. At other times the original proof is so beautiful or provide a vital lessons
in mathematical reasoning that I let it's author speak directly (such as Euclid's proof of the Pyteagorian theorem).

The knowledge and proofs contained herein is taken from a multitude of sources, many have been rewritten or reordered so to make
them more accessible, but often one finds mathematics written with such breathtaking clarity as so to render any further attempt
upon simplification impossible. In these cases the information is simply restated and put in context. In the end of each chapter
propper credit is provided as well as suggestions on further reading for the adventures person.\\ 
\\
\indent Mathematics is utility and its usage is spread far and wide; from logical reason about programming languages to the physical sciences. So stories about its usages is included as separate capsules whenever I have found it fitting. At the same time mathematics is also history, from Archimedes war machines used against the intruding romans to everyday stories such as Newton's remark that he "do not like to be teased by foreigners about mathematical things". In the end all that I hope to achieve is to shine a little light on the fascinating field of human creativity known as mathematics, to hopefully illustrate the meaning of the the great philosopher Spinoza's words, when he said that "God is a mathematician". \\
\flushright Lars Tackmann \\ Copenhagen, Denmark\flushleft
